How to publish in Biological Sciences: a guide for the uninitiated

### 2021-12-15 4.07-retract.Rmd
Add files via upload
# with lots of changes following copy edits and proofs

| Action | Example | Conclusions impacted | Issued by |
| - | - | - | - |
| Corrigendum / erratum / correction | Important typos / incorrect figure legends or tables / author name or address issues | Not | Author |
| Expression of concern | Data appear unreliable / misconduct suspected | Undetermined | Editor (perhaps through information received) |
| Partial retraction | One aspect of the study is corrupted / Inappropriate analysis | Overall findings remain | Author or Editor |
| Full retraction | Majority of the study is corrupted / Evidence of misconduct / Work is invalidated | Yes | Author or Editor |

@@

An expression of concern lies between a correction and a retraction. An expression of concern can precede a retraction, but suggests that the editors are seriously worried about something being wrong with the publication. For example, a paper published in _Proceedings of the National Academy of Sciences_ that used an unusual mutant strain of _Chlamydomonas_ (a genus of green algae) was placed under an official 'Editorial Expression of Concern' when the editors learned that the authors would not share their strain with any other researchers [@national2021editorial]. Clearly, this notice could be removed if, for example, the authors agreed to share their strain and their findings were replicated. However, if they continue to refuse, the editors could also fully retract the paper.

@@

Although retractions are so rare in life sciences, 0.06% of all papers published between 1990 and 2020, they appear to be on the increase in the last 30 years: from around 0.5 to over 15 papers per 10 000 published (see Figure \@ref(fig:life-sciences-retractions)). It takes a mean time of nearly 2 years between notification of problems with a paper, and issuing a correction or retraction [@grey2021timeliness], but this belies a bimodal distribution in retraction times with the first hump coming from self-correcting authors, or clerical errors from journals coming within months of the original date of publication. The second hump is usually associated with [fraud](#preittdata), and comes after several years of investigations by institutions often with added legal frustrations.

(ref:life-sciences-retractions) **The growth of retractions in life sciences journals over time.** The three lines show when the original paper is published (blue line) and when a correction, expression of concern or retraction are made (black line). Articles that are finally retracted (green line) are only a part of those with other issues after 2008. While the number of papers published that are later retracted appears to take turn downward from 2014, this may simply represent the lag before they are later retracted. This data came from the Retraction Watch database, selecting only data from  Basic Life Sciences and Environment. The data is normalised by dividing the total number of publications (as taken from SCOPUS) in this area for the year by 10 000, and multiplying this by numbers of Original or Retracted papers.

```{r life-sciences-retractions, echo=FALSE, fig.cap="(ref:life-sciences-retractions)"}
if(knitr::is_html_output(excludes="markdown")) knitr::include_url("https://johnmeasey.github.io/Life-sciences-retractions/index.html", height="550px") else knitr::include_graphics("figures/life-sciences-retractions/life-sciences-retractions.png")
```

Retractions showed a steep uptake in 2011, when a number of laboratories that made multiple retractions in 2011 [see @fang2011retracted; @brainard2018what]. A single publisher was responsible for pulling a great many of the retracted papers in 2011, and this spike in retractions isn't seen in the life sciences [@oransky2011year; @brainard2018what]. What is clear from Figure \@ref(fig:life-sciences-retractions) is that the rising trend in retractions (black line) appears to have been unaffected by the 2011 spike in other subject areas. Claims that retractions are levelling off [@brainard2018what], are not matched by the data.

@@

Publishers are bad at sending through the correct metadata with their content. For example, for the same 30 year period as Figure \@ref(fig:life-sciences-retractions), Web of Science lists only 133 retractions. Really what this means that if you aren't using Zotero, and you want to be sure that there hasn't been a retraction in any of your source material, you need to run a search on the [Retraction Watch database](https://retractionwatch.com) - better to use Zotero says Ivan Oransky. Right now, the chance that any paper published in the last 30 years that you have cited will get retracted is still low (one in 1750), but if it was published in 2020 this rises to one in 650.

@@

@@

But we must be cautious about playing a blame game when it comes to journal retractions [@smith2021reimagining]. We already know that peer review has shortcomings (see [Part 4](#problempeerreview)), and even the best of peer reviewers and/or editors cannot be expected to spot potentially fatal errors, especially when these come about from deliberate deceit on the part of the authors. Retraction will remain a necessary part of the scientific publishing process, and as retractions become more commonplace among journals, we can hope that guidelines will be followed in a timely manor [@grey2021timeliness].

@@

## Fabrication of data {#fabrication}
The fabrication of data does happen. A growing body of retractions and alleged evidence on the tampering of data in spreadsheets has led to the suspension of a top Canadian researcher, [Jonathan Pruitt](#Pruittdata). The detection of fraudulent (usually duplicated) data in spreadsheets is not too difficult to spot (e.g. by application of Benford's law - the last digits of naturally occurring numbers should approach uniformity), and has become the subject of some contract data scientists who specialise in finding such instances of fraud. To some extent, the automated assessment of fraudulent practices has been or could be implemented for many infringements [@bordewijk2021methods].

@@

Some have suggested that the pressure to obtain a permanent academic position is enough to drive some scientists to commit fraud [@husemann2017publicationism; @kun2018publish; @fanelli2015misconduct]. The idiom ‘[publish or perish](#publishmore)’, and the importance of publishing is made elsewhere. However, I hope that by shedding some light on unethical practices, this book equips you to avoid these together with those that may espouse them, and instead show you that there is a better path to success.

[Pruitt’s case](#PruittData) highlights a good reason for increased transparency in the publication process. A blog post from someone caught up in the Pruitt retractions makes the point that journals that insisted on full data deposits for publication were well ahead of those that hadn't [@bolnick202117]. Of growing concern in many areas of Biological Sciences is the potential to manipulate results that are essentially images of results, for example blots on a gel. However, it turns out that manipulated images are also not too hard to discern.

### Who is responsible?
In the case of fraud, retraction statements should indicate who the perpetrator is in order to exonerate the other researchers. Some research into the likely source of the fraud has been conducted. There are clearly [serial fraudsters](#PruittData), and the presence of their names in the author list is a red flag for those investigating fraud. Data from papers that are known to be fraudulent suggest that the first author is the most likely to be responsible for the fraud committed, and middle authors the least [@hussinger2019scientific]. This suggests that in a collaboration, you should be very careful who you collaborate with. While you might not be responsible, the discovery of (particularly large scale) fraud might well harm your career.

In a study looking for patterns about types of authors involved in retractions (all reasons), suggested that Early Career Researchers were particularly likely to be involved in retractions [@fanelli2015misconduct], although the exact reason why remains obscure.

It is worth noting that while the journals (and ultimately the [journal editor](#gatekeepers)) are responsible for retractions from journals, this is not the same as punishing individuals who have committed fraud. As we have already seen, there can be many reasons for a retraction, and it is not up to editors or journals to punish researchers as there will be innocent researchers who may also need to make retractions. Moreover, it should never be the role of the journal to have any punitive action over a researcher. There are other mechanisms for this with the employer and (where applicable) the academic society involved. Different institutions and governments will have different rules when it comes to fraud being committed by their employees. These processes may be legal and take some time to finish. As we saw with the case of [Pruitt](#Pruittdata), once lawyers get involved, the process may become mired in a lot more beaurocracy.

@@

## What to do if you suspect others {#fraud}
If you suspect that someone in a lab in your department, faculty or university is fabricating data, find out whether your university has a research integrity officer (RIO); most universities in the US have one. Document your evidence if you can and approach the RIO or person in the equivalent position. If you can't find such a person, then ask at you university library for the most relevant person. Libraries are usually neutral places where you can find out information without arousing suspicion. You do need to be careful that you do not place yourself in harm's way when reporting, so be prudent about sharing until you are assured protection from any potential retaliation. It isn't easy to be a whistleblower - but it is the right thing to do.

@@

## Confirmation Bias and the paradox of high-flying academic careers {#PruittData}
Jonathan Pruitt had it all going for him. His studies of spider sociality were producing significant results that opened the door to publications in high impact journals. In turn, this opened the door to getting prizes and funding. The funding allowed him to conduct more studies and soon a prestigious chair in Canada with more funding to pursue his rocketing career. Things started to unravel for Pruitt when colleagues raised concern about the data in some of his publications. Things gathered pace very quickly, and doubt gathered around more and more of his publications. Although there is much written about the Pruitt debacle on the internet, the blog by Am. Nat. editor and former Pruitt fan and friend, Dan Bolnick, is particularly enlightening [@bolnick202117]. Pruitt’s case is becoming increasingly untenable as more editors backed by co-authors are retracting papers where he contributed data [see @marcus2020spider]. For Pruitt, this has become a threat to his career and livelihood [@pennisi2020embattled]. Similarly, his university is facing the possibility that they hired a fraud. Consequently, this whole debacle has slipped into the legal world. Bolnick has clearly suffered personally from the affair, but has set out to provide as transparent an account as possible. Of note are his assurances to one of Pruitt's co-authors to:

> "...use transparency to gain community support for the retraction process." 
>
> --- @bolnick202117

The harrowing part of Bolnick's account is when he, co-authors and other editors started to receive letters from Pruitt's Lawyer (see one example on [Bolnick's blog](https://ecoevoevoeco.blogspot.com/2021/05/17-months.html)). At the point that the lawyer steps in, the functioning academic community that had raised itself to meet the demands of the concerns began to get muted in what Bolnick calls a 'chilling effect'. Bolnick then makes an important point that the legal threats from Pruitt's Lawyer were stifling the freedom for academics (in this case the co-authors and editors) to publish.

@@

It is clear from these reports that there are systemic problems when high profile scientists are accused of fraud. Journals say that it's the responsibility of the institutions, and the institutions have no impetus to find fraud as that might lose them a very productive (think research income) and high profile scientist. What university would want to have its name dragged through the mud, and on top of this lose a large amount of grant income? Top researchers become untouchables in many institutions because they are essentially cash cows that no-one wants to disturb. Allegations against such individuals also include [bullying](#bullying) and sexual misconduct. For those interested in reading more high profile misdemeaners in science Stuart Ritchie [-@ritchie2020science] has put together a popular book on the subject.

>>>>

| Action | Example | Conclusions impacted | Issued by |
| --- | ------- | --- | --- |
| Corrigendum / erratum / correction | Important typos / incorrect figure legends or tables / author name or address issues | Not | Author |
| Expression of concern | Data appear unreliable / misconduct suspected | Undetermined | Editor (perhaps through information received) |
| Partial retraction | One aspect of the study is corrupted / Inappropriate analysis | Overall findings remain | Author or Editor |
| Full retraction | Majority of the study is corrupted / Evidence of misconduct / Work is invalidated | Yes | Author or Editor |

@@

An expression of concern lies between a correction and a retraction. An expression of concern can precede a retraction, but suggests that the editors are seriously worried about something being wrong with the publication. For example, a paper published in _Proceedings of the National Academy of Sciences_ that used an unusual mutant strain of _Chlamydomonas_ (a genus of green algae) was placed under an official 'Editorial Expression of Concern' when the editors learned that the authors would not share their strain with any other researchers [@berenbaum2021editorial]. Clearly, this notice could be removed if, for example, the authors agreed to share their strain and their findings were replicated. However, if they continue to refuse, the editors could also fully retract the paper.

@@

Although retractions are so rare in life sciences, 0.06% of all papers published between 1990 and 2020, they appear to be on the increase in the last 30 years: from around 0.5 to over 15 papers per 10 000 published (see Figure \@ref(fig:life-sciences-retractions)). It takes a mean time of nearly 2 years between notification of problems with a paper, and issuing a correction or retraction [@grey2021timeliness], but this belies a bimodal distribution in retraction times with the first hump coming from self-correcting authors, or clerical errors from journals coming within months of the original date of publication. The second hump is usually associated with [fraud](#pruittdata), and comes after several years of investigations by institutions often with added legal frustrations.

(ref:life-sciences-retractions) **The growth of retractions in life sciences journals over time.** The three lines show when the original paper is published (blue line) and when a correction, expression of concern or retraction are made (black line). Articles that are finally retracted (green line) are only a part of those with other issues after 2008. While the number of papers published that are later retracted appears to take a turn downward from 2014, this may simply represent the lag before they are later retracted. This data came from the Retraction Watch database, selecting only data from  Basic Life Sciences and Environment. The data is normalised by dividing the total number of publications (as taken from SCOPUS) in this area for the year by 10 000, and multiplying this by numbers of Original or Retracted papers.

```{r life-sciences-retractions, echo=FALSE, fig.cap="(ref:life-sciences-retractions)"}
if(knitr::is_html_output(excludes="markdown")) knitr::include_url("https://johnmeasey.github.io/Life-sciences-retractions/index.html", height="450px") else knitr::include_graphics("figures/life-sciences-retractions/life-sciences-retractions.png")
```

Retractions showed a steep uptake in 2011, when a number of laboratories made multiple retractions [see @fang2011retracted; @brainard2018what]. A single publisher was responsible for pulling a great many of the retracted papers in 2011, and this spike in retractions isn't seen in the life sciences [@oransky2011year; @brainard2018what]. What is clear from Figure \@ref(fig:life-sciences-retractions) is that the rising trend in retractions (solid black line) appears to have been unaffected by the 2011 spike in other subject areas. Claims that retractions are levelling off [@brainard2018what], are not matched by the data.

@@

@@

Publishers are bad at sending through the correct metadata with their content. For example, for the same 30 year period as Figure \@ref(fig:life-sciences-retractions), Web of Science lists only 133 retractions. Really what this means is that if you aren't using Zotero, and you want to be sure that there hasn't been a retraction in any of your source material, you need to run a search on the [Retraction Watch database](https://retractionwatch.com) - "better to use Zotero", says Ivan Oransky. Right now, the chance that any paper published in the last 30 years that you have cited will get retracted is still low (one in 1750), but if it was published in 2020 this rises to one in 650.

@@

But we must be cautious about playing a blame game when it comes to journal retractions [@smith2021reimagining]. We already know that peer review has shortcomings (see [Part IV](#problempeerreview)), and even the best of peer reviewers and/or editors cannot be expected to spot potentially fatal errors, especially when these come about from deliberate deceit on the part of the authors. Retraction will remain a necessary part of the scientific publishing process, and as retractions become more commonplace among journals, we can hope that guidelines will be followed in a timely manor [@grey2021timeliness].

@@

## Fabrication of data {#fabrication}

The fabrication of sensational results that might land you publication in a high ranking journal, and consequently a job or a promotion, is an important incentive for individuals to [commit fraud](#PruittData). However, there are also a large number of people who need a publication in any international journal but lack the scientific or linguistic skills to make it happen. These people are more likely to pay third parties of organised fraudsters to have their name appear on an author line. The resulting paper in which all text as well as the data is fabricated, represents another source of large numbers of retractions in lower ranking journals.

### Fraud by individuals or laboratories
The fabrication of data does happen. An anonymous survey on research integrity in the Netherlands suggested that prevalence of fabrication was 4.3% and falsification was 4.2%, while nearly half of those surveyed admitted to questionable research practices most prevalent in PhD candidates and junior researchers [@gopalakrishna2021prevalence]. A growing body of retractions and alleged evidence on the tampering of data in spreadsheets has led to the suspension of a top Canadian researcher, [Jonathan Pruitt](#Pruittdata). The detection of fraudulent (usually duplicated) data in spreadsheets is not too difficult to spot (e.g. by application of Benford's law - the last digits of naturally occurring numbers should approach uniformity), and has become the subject of some contract data scientists who specialise in finding such instances of fraud. To some extent, the automated assessment of fraudulent practices has been or could be implemented for many infringements [@bordewijk2021methods].

[Pruitt’s case](#PruittData) (see below) highlights a good reason for increased transparency in the publication process. A blog post from someone caught up in the Pruitt retractions makes the point that journals that insisted on full data deposits for publication were well ahead of those that hadn't [@bolnick202117]. Of growing concern in many areas of Biological Sciences is the potential to manipulate results that are essentially images of results, for example blots on a gel. However, it turns out that manipulated images are also not too hard to discern.

### By third parties

@@

The pressure to publish is widely acknowledged as driving questionable research practices, including fraud [@gopalakrishna2021prevalence]. Some have suggested that the additional pressure to obtain a permanent academic position is enough to drive some scientists to commit fraud [@husemann2017publicationism; @kun2018publish; @fanelli2015misconduct]. The idiom ‘[publish or perish](#publishmore)’, and the importance of publishing is made elsewhere. However, I hope that by shedding some light on unethical practices, this book equips you to avoid these together with those that may espouse them, and instead show you that there is a better path to success.

When people rely on publishing papers in journals for their careers, it seems that there are predators happy to make a publication happen at the expense of the integrity of the publishing process. Scammers posing as well-known academics use modified domain names to gain editorial control of [special issues](#symposium) of journals. Once in the editorial seat, they quickly start accepting low-quality papers from clients who pay to get included [@else2021scammers]. Retractions of hundreds of papers in [special issues](#symposium) indicate that editors need to pay special attention before handing over the role of [gatekeeper](#boards).

## Who is responsible?
In the case of fraud, retraction statements should indicate who the perpetrator is in order to exonerate the other researchers. Some research into the likely source of the fraud has been conducted. There are clearly [serial fraudsters](#PruittData), and the presence of their names in the author list is a red flag for those investigating fraud. Data from papers that are known to be fraudulent suggest that the first author is the most likely to be responsible for the fraud committed, and middle authors the least [@hussinger2019scientific]. This suggests that in a collaboration, you should be very careful who you collaborate with. While you might not be responsible, the discovery of (particularly large scale) fraud might well harm your career.

A study looking for patterns about types of authors involved in retractions (all reasons), suggested that Early Career Researchers were particularly likely to be involved in retractions [@fanelli2015misconduct], although the exact reason why remains obscure.

It is worth noting that while the journals (and ultimately the [journal editor](#gatekeepers)) are responsible for retractions from journals, this is not the same as punishing individuals who have committed fraud. As we have already seen, there can be many reasons for a retraction, and it is not up to editors or journals to punish researchers as there will be innocent researchers who may also need to make retractions. Moreover, it should never be the role of the journal to have any punitive action over a researcher. There are other mechanisms for this with the employer and (where applicable) the academic society involved. Different institutions and governments will have different rules when it comes to fraud being committed by their employees. These processes may be legal and take some time to finish. As we will see with the case of [Pruitt](#Pruittdata), once lawyers get involved, the process may become mired in a lot more bureaucracy.

@@

## What to do if you suspect others {#fraud}
If you suspect that someone in a lab in your department, faculty or university is fabricating data, find out whether your university has a research integrity officer (RIO); most universities in the US have one. Document your evidence if you can and approach the RIO or person in the equivalent position. If you can't find such a person, then ask at your university library for the most relevant person. Libraries are usually neutral places where you can find out information without arousing suspicion. You do need to be careful that you do not place yourself in harm's way when reporting, so be prudent about sharing until you are assured protection from any potential retaliation. It isn't easy to be a whistleblower - but it is the right thing to do.

@@

## Confirmation Bias and the paradox of high-flying academic careers {#PruittData}
Jonathan Pruitt had it all going for him. His studies of spider sociality were producing novel and significant results that opened the door to publications in high impact journals. In turn, this opened the door to getting prizes and funding. The funding allowed him to conduct more studies and soon a prestigious chair in Canada with more funding to pursue his rocketing career. Things started to unravel for Pruitt when colleagues raised concern about the data in some of his publications. Things gathered pace very quickly, and doubt gathered around more and more of his publications. Although there is much written about the Pruitt debacle on the internet, the blog by *American Naturalist* editor and former Pruitt fan and friend, Dan Bolnick, is particularly enlightening [@bolnick202117]. Pruitt's case is becoming increasingly untenable as more editors backed by co-authors are retracting papers where he contributed data [see @marcus2020spider]. For Pruitt, this has become a threat to his career and livelihood [@pennisi2020embattled]. Similarly, his university is facing the possibility that they hired a fraud. Consequently, this whole debacle has slipped into the legal world. Bolnick has clearly suffered personally from the affair, but has set out to provide as transparent an account as possible.

The harrowing part of Bolnick's account is when he, co-authors and other editors started to receive letters from Pruitt's Lawyer (see one example on [Bolnick's blog](https://ecoevoevoeco.blogspot.com/2021/05/17-months.html)). At the point that the lawyer steps in, the functioning academic community that had raised itself to meet the demands of the concerns began to get muted. Bolnick then makes an important point that the legal threats from Pruitt's Lawyer were stifling the freedom for academics (in this case the co-authors and editors) to publish, and therefore their [academic freedom](#tenure).

@@

It is clear from these reports that there are systemic problems when high profile scientists are accused of fraud. Journals say that it's the responsibility of the institutions, and the institutions have no impetus to find fraud as that might lose a very productive (think research income) and high profile scientist [see @pennisi2021data]. What university would want to have its name dragged through the mud, and on top of this lose a large amount of grant income? Top researchers become untouchables in many institutions because they are essentially cash cows that no-one wants to disturb. Allegations against such individuals also include [bullying](#bullying) and sexual misconduct. For those interested in reading more high profile misdemeaners in science Stuart Ritchie [-@ritchie2020science] has put together a popular book on the subject.

### 2021-12-15 4.08-bullying.Rmd
Add files via upload
# with lots of changes following copy edits and proofs

I am going to talk about academic bullying because it is currently prevalent in academia, and because most of the bullies are unaware of what they are doing. In a 2019 survey of graduate students (see [here](https://www.nature.com/articles/d41586-019-03459-7) Figure \@ref(fig:bully)), 22% felt that they had experienced bullying during their PhD program. The only way of improving the situation around academic bullying is for everyone to become more aware. It may not be happening to you, but it may be happening to people around you either in your lab or in another lab in the same department or faculty. If you think that this is very rare behaviour in academia, think again [@devlin2018hundreds].

(ref:bully) Responses to a survey of graduate students (in 2019) demonstrates the changes in bullying propensity in different research cultures.

@@

## How to spot a bully
A bully is anyone who abuses or misuses their position of power in order to humiliate, denigrate or injure another. This does not need to be your advisor, or someone in your lab. It can be anyone in your working environment (watch the video [here](https://youtu.be/bZmmp7i9Tsc) or [here](https://youtu.be/lKUONMm-pWo)). These people are usually in positions of power with influence over you and your future. As in my example (below) you may be worried that the power they have could be used by them to negatively impact your future. If you are worried about this, then their behaviour most likely conforms to bullying. Bullying often involves harassment that is designed to undermine your dignity, often through sexism, racism, or another prejudice [@krishna2020playing]. Even if you think that your bully didn’t mean to cause offence, the fact that they did upset you and that this behaviour was unwanted is enough to fulfil the criteria for bullying. Thus, it is not what they intended, but what you felt that is important in bullying. A direct consequence of this is that bullies often don’t recognise this as a description of their behaviour. It is worth bearing in mind that bullies are often damaged individuals who are repeating behaviour that they have themselves experienced from others. This doesn’t excuse their behaviour, but they may think that such behaviour is normal.

@@

Avoid the bully when asking for your institutional rules, but you should be able to find them via your departmental secretary, administrative support staff in the department or faculty. Read the documentation carefully and learn about how and by whom such reports are dealt with. Become aware of resources that are available; [paritymovement.org](https://paritymovement.org/) is a great place to start. Read more about other people’s experiences and be aware that you are not alone [@mahmoudi2020survivor, @malaga2004meyer].

Next, document your case. Make some notes about the incident(s), when they happened, what was said, how you were made to feel and what power you feel the person has over you.

@@

Follow your university’s rules about who to go to with your complaint. Don’t leave it to the next person in your lab to experience, they may not be as strong or as resourceful as you. It may not be your career that is destroyed, but the next student might not be so lucky.

>>>>

I am going to write here about academic bullying because it is currently prevalent in academia, and because most of the bullies are unaware of what they are doing. In a 2019 survey of graduate students (see [here](https://www.nature.com/articles/d41586-019-03459-7) Figure \@ref(fig:bully)), 22% felt that they had experienced bullying during their PhD program. The only way of improving the situation around academic bullying is for everyone to become more aware. It may not be happening to you, but it may be happening to people around you either in your lab or in another lab in the same department or faculty. If you think that this is very rare behaviour in academia, think again [@devlin2018hundreds].

(ref:bully) **Do you feel that you have experienced bullying in your PhD program?** Responses to a survey of graduate students (in 2019) demonstrates the changes in bullying propensity in different research cultures.

@@

## How to spot a bully
A bully is anyone who abuses or misuses their position of power in order to humiliate, denigrate or injure another. This does not need to be your advisor, or someone in your lab. It can be anyone in your working environment (watch the video [here](https://youtu.be/bZmmp7i9Tsc) or [here](https://youtu.be/lKUONMm-pWo)), including people in your institutes' administration. These people are usually in positions of power with influence over you and your future. As in my example (below) you may be worried that the power they have could be used by them to negatively impact your future. If you are worried about this, then their behaviour most likely conforms to bullying. Bullying often involves harassment that is designed to undermine your dignity, often through sexism, racism, or another prejudice [@krishna2020playing]. Even if you think that your bully didn’t mean to cause offence, the fact that they did upset you and that this behaviour was unwanted is enough to fulfil the criteria for bullying. Thus, it is not what they intended, but what you felt that is important in bullying. A direct consequence of this is that bullies often don’t recognise this as a description of their behaviour. It is worth bearing in mind that bullies are often damaged individuals who are repeating behaviour that they have themselves experienced from others. This doesn’t excuse their behaviour, but they may think that such behaviour is normal.

@@

Avoid the bully when asking for your institutional rules, but you should be able to find them via your departmental secretary, administrative support staff in the department or faculty. Read the documentation carefully and learn about how and by whom such reports are dealt with. Become aware of resources that are available; [paritymovement.org](https://paritymovement.org/) is a great place to start. Read more about other people’s experiences and be aware that you are not alone [@mahmoudi2020survivor; @malaga2004meyer].

Next, document your case. Make some notes about the incident(s), when they happened, what was said, how you were made to feel, and what power you feel the person has over you.

@@

Follow your university’s rules about who to go to with your complaint. Don’t leave it to the next person in your lab to experience, they may not be as strong or as resourceful as you. It may not be your career that is destroyed, and the next student might not be so lucky.

### 2021-12-15 5.00-lastnote.Rmd
Add files via upload
# with lots of changes following copy edits and proofs

I really hope that this book has been helpful and that it has achieved what I set out to do: provide you with the guide on how to publish in biological sciences. If you feel that this book has important items missing, is out-of-date or simply wrong, then please [contribute](#contribute). Any good guide relies upon the people that use it to keep it viable. My special plea to those of you who are Early Career Scientists is to become part of the solution for shifting from the [closed to open models](#openscience) of scientific publishing. In doing so, you will make science more transparent, open and equitable. While there is much that is currently wrong in the system, the vast majority of the actors in it are working for the good of science. There are some simple steps to achieving this.

As scientists, we need to go back to our core concerns: **the scientific method, independence, transparency and reproducibility**. Join your disciplinary [academic society](#societies) and stand for election to a position where your voice will be heard. Join the [gatekeepers](#gatekeepers) of the society's journals (if you have time to read), and use your soft power to help bring about change.

1. Make more use of archiving services: [preprints](#preprints2), data and proposals
1. Make use of independent peer review services: peer community in, and  peer commons
1. Persuade your society to move to a [Diamond Open Access](#diamond) [overlay](#overlay) journal model

If I could emphasise one really important aspect of all of your Early Career; that is to try to enjoy it!

>>>>

I really hope that this book has been helpful and that it has achieved what I set out to do: provide you with the guide on how to publish in biological sciences. If you feel that this book has important items missing, is out-of-date or simply wrong, then please [contribute](#contribute). Any good guide relies upon the people that use it to keep it viable. My special plea to those of you who are Early Career Scientists is to become part of the solution for shifting from the [closed to open models](#openscience) of scientific publishing. In doing so, you will make science more transparent, open and equitable. While there is much that is currently wrong in the system, the vast majority of the actors in it are working for the good of science. While I conceded that publishing science is a [wicked problem](#wicked) which is not easily solved, there are some simple steps that we can all make toward a resolution that benefits science.

When seeking a solution to publishing science, and publishing in other academic disciplines, we cannot compromise on the cornerstones of the scientific method: **rigour, independence, transparency and reproducibility**. This book should have highlighted aspects of scientific publishing that currently violate these cornerstones and so need to be changed. When you can, select those journals and publishers that maintain the [highest standards](#transparent). When you can't, be aware that the [motives of many big publishers](#capture) are not toward [transparency and openness](#openness). Join your discipline's [academic society](#societies) and stand for election to a position. Issue your clarion call for change. Join the [gatekeepers](#gatekeepers) of the society's journals, and use your soft power to help bring about change.

1. Make more use of archiving services: [preprints](#preprints2), data and proposals
1. Make use of independent peer review services: peer community in, and  peer commons
1. Persuade your society to move to a [Diamond Open Access](#diamond) [overlay](#overlay) journal model
1. Become part of the solution, start an [Open Science Community](osf.io/vz2sy/) in your institution

### 2021-12-15 index.Rmd
Add files via upload
# with lots of changes following copy edits and proofs

--- 
title: "How to publish in Biological Sciences: a guide for the uninitiated"
author: "John Measey"
date: "`r Sys.Date()`"
documentclass: krantz
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
colorlinks: yes
lot: yes
lof: yes
site: bookdown::bookdown_site
output: bookdown::gitbook
description: "Publishing a paper in an academic journal should simply consist of submitting a publication worthy manuscript. But having a working knowledge of publishing will enable you to make better decisions about what, where and how to submit manuscripts. This all comes with experience, and in this book, I try to explore the areas of assumed knowledge, and furnish them with explanations pitched at the Early Career Researcher, along with links and citations where you can read more. I explain the many choices that exist for those wishing to submit a manuscript for publication in the Biological Sciences. I explore the world of publication bias, and how this is evidenced by reviewers and editors. Impact Factors, citations and how the desire to track the performance of academics has led to unethical practices and exploitation of science and scientists. This guide provides an 'everything you wanted to know about publishing but were afraid to ask' approach for anyone who still feels that submitting a manuscript is like posting it into a black box. This book is written to get you onto an even footing."
github-repo: yihui/bookdown-crc
graphics: yes
cover-image: images/Cover.png
favicon: images/favicon.png
---

@@

# Under Contract  {-}

This book is currently under contract with CRC Press. This version was **submitted** for publication on: __27 August 2021__

@@

@@

There are a great many people that I need to thank. First and foremost are the members of my lab, past and present, who have inspired me to put together first the blog posts and then the book. It is because you wanted more that I put this together. This book contains lots of links to blogs and articles written and posted freely on the internet by others who also aim to demystify and help. I thank this greater academic community (especially \#academicTwitter) for sharing and inspiring. Thanks go to the many reviewers and editors who have taken their time to improve my writing. I am still learning. Lots of the text in this book has been improved by feedback from my students and postdocs. A special mention must go to my brother, Richard, who has hosted my lab website for more than a decade, and especially for saving blog posts from hacking attacks. Thanks also to my wife, Thalassa, who proofread many of the blog posts after I had published them late at night, so that I could correct them over breakfast in the morning. James Baxter-Gilbert, Dan Bolnick, Jack Dougherty, Anthony Herrel, Michael Hochberg, Allan Ellis, Rachael Lammey (CrossRef), Andrea Melotto, Lisa and Mark O'Connell, Ivan Oransky (Retraction Watch), Heather Piwowar (ImpactStory), Claire Riss (Center for Open Science), Johan Rooryck (cOAltition S), James Vonesh, Carla Wagener all read or commented on different aspects of the book. Thanks are also due to my colleagues at the Centre for Invasion Biology, the Department of Botany and Zoology, and Stellenbosch University. A special thanks to the librarians who have supported many of my more extreme rantings about publishers.

>>>>

--- 
title: "How to publish in Biological Sciences: a guide for the uninitiated"
author: "John Measey"
date: "`r Sys.Date()`"
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
csl: peerj.csl
link-citations: yes
colorlinks: yes
lot: yes
lof: yes
site: bookdown::bookdown_site
output: bookdown::gitbook
description: "Publishing a paper in an academic journal should simply consist of submitting a publication worthy manuscript. But having a working knowledge of publishing will enable you to make better decisions about what, where and how to submit manuscripts. This all comes with experience, and in this book, I unravel areas of assumed knowledge and furnish them with explanations pitched at the Early Career Researcher, along with links and citations where you can read more. This book explores the wicked problem of academic publishing, and how to move from a closed to an open, transparent system. I explain the many choices that exist for those wishing to submit a manuscript for publication in the Biological Sciences. I explore the world of publication bias, and how this is evidenced by reviewers and editors. Impact Factors, citations and how the desire to track the performance of academics has led to unethical practices and exploitation of science and scientists. This guide provides an 'everything you wanted to know about publishing but were afraid to ask' approach for anyone who still feels that submitting a manuscript is like posting it into a black box. This book is written to get you onto an even footing."
github-repo: yihui/bookdown-crc
graphics: yes
cover-image: images/Cover.png
favicon: images/favicon.png
---

@@

# In Press  {-}

This book is currently **in press** with CRC Press. This version follows proofs returned on: __15 December 2021__. The book is expected to be available on 15 March 2022.

@@

You can pre-order your own copy on the [CRC Press website](https://www.routledge.com/How-to-Publish-in-Biological-Sciences-A-Guide-for-the-Uninitiated/Measey/p/book/9781032116419), [Amazon](https://www.amazon.com/How-Publish-Biological-Sciences-Uninitiated/dp/1032116447), [Barnes & Noble](https://www.barnesandnoble.com/w/how-to-publish-in-biological-sciences-john-measey/1140172276) and other good book sellers.

@@

There are a great many people that I need to thank. First and foremost are the members of my lab, past and present, who have inspired me to put together first the blog posts and then the book. It is because you wanted more that I put this together. This book contains lots of links to blogs and articles written and posted freely on the internet by others who also aim to demystify and help. I thank this greater academic community (especially \#academicTwitter) for sharing and inspiring. Thanks go to the many reviewers and editors who have taken their time to improve my writing. I am still learning. Lots of the text in this book has been improved by feedback from my students and postdocs. A special mention must go to my brother, Richard, who has hosted my lab website for more than a decade, and especially for saving blog posts from hacking attacks. Thanks also to my wife, Thalassa, who proofread many of the blog posts after I had published them late at night, so that I could correct them over breakfast in the morning. Graham Alexander, James Baxter-Gilbert, Dan Bolnick, Jack Dougherty, João Fontiela, Anthony Herrel, Michael Hochberg, Allan Ellis, Rachael Lammey (CrossRef), Andrea Melotto, Lisa and Mark O'Connell, Ivan Oransky (Retraction Watch), Heather Piwowar (ImpactStory), Claire Riss (Center for Open Science), Johan Rooryck (cOAltition S), Daniel van Blerk, James Vonesh, Carla Wagener all read or commented on different aspects of the book. Thanks are also due to my colleagues at the Centre for Invasion Biology, the Department of Botany and Zoology, and Stellenbosch University. A special thanks to the librarians who have supported many of my more extreme rantings about publishers.
